---
title: "Comparing Linear Models for SDG 14: Trawling in the Latin America and Caribbean region"
author: "Brandon Koskie"
format: 
  revealjs:
    theme: simple
    slide-number: true
    transition: fade
    scrollable: true
    toc: true
    toc-depth: 2
editor: visual
---

## Introduction to Latin America and the Caribbean

-   🐟 Fisheries in Latin America and the Caribbean

    -   🌊 **LAC is one of the world's richest marine biodiversity regions**, with vast coastlines, coral reefs, and exclusive economic zones.

    -   🐠 **Fisheries support millions of livelihoods**, especially in small island developing states and coastal communities.

    -   🍽️ **Overfishing and illegal, unreported, and unregulated (IUU) fishing** threaten marine ecosystems and food security.

    -   ⚖️ **Trawling is common** in the region but is often unsustainable, contributing to bycatch and habitat degradation.

    -   📉 **Fish stocks in many countries are declining**, with several species listed as overexploited.

    -   🛡️ **Marine protected areas (MPAs)** are expanding but vary in enforcement and effectiveness across countries.

    -   📊 **Monitoring and data collection are limited**, making it difficult to evaluate the impact of conservation efforts.

-   SDG 14: Life Below Water

    ## Why This Matters

    -   **Overfishing** and **marine ecosystem degradation** threaten food security, tourism, and livelihoods.
    -   Understanding key predictors of SDG 14 can help target policy and conservation efforts.
    -   Regression modeling helps identify high-impact factors.\
        \

------------------------------------------------------------------------

# 

## Research Question

**Does adding more indicators improve predictive performance of trawling activity?**

We compare two models:

-   **Model 1**: Marine Protected Areas + Discarded Fish\
-   **Model 2**: Model 1 + Fish Stocks + Clean Water

------------------------------------------------------------------------

# Data and Variables

-   Source: 2023 Sustainable Development Report (SDR)
-   Country-level, normalized indicators (0–100 scale)
-   **Target Variable**: SDG 14 Trawling Indicator\
-   **Predictors**:
    -   Marine Protected Areas
    -   Discarded Fish
    -   Fish Stocks
    -   Coastal Population

------------------------------------------------------------------------

```{r}

library(tidyverse)

# Load your dataset
data <- read_csv("SDR-2023-Data.csv")

# Filter for Caribbean countries using the region column
LAC_data <- data %>%
  filter(`Regions used for the SDR` == "LAC")


```

```{r}
# Select relevant columns and rename them
LAC_data_renamed <- LAC_data %>%
  select(
    `Country`, 
    `normalized_score_sdg14_trawl` = `Normalized Score: sdg14_trawl`,
    `normalized_score_sdg14_discard` = `Normalized Score: sdg14_discard`,
    `normalized_score_sdg14_cpma` = `Normalized Score: sdg14_cpma`,
    `normalized_score_sdg14_fishstocks` = `Normalized Score: sdg14_fishstocks`,
    `normalized_score_sdg14_cleanwat` = `Normalized Score: sdg14_cleanwat`
  )

```

```{r}
# Preview the first few rows of the renamed data
head(LAC_data_renamed)
```

## Models

```{r}
# Remove rows with any missing values
LAC_data_clean <- LAC_data_renamed %>% drop_na()

# Model 1: Simple model with only discarded fish and marine protected areas
m1 <- lm(normalized_score_sdg14_trawl ~ normalized_score_sdg14_discard + normalized_score_sdg14_cpma,
         data = LAC_data_clean)
summary(m1)

# Model 2: Add fish stocks and clean water
m2 <- lm(normalized_score_sdg14_trawl ~ normalized_score_sdg14_discard + normalized_score_sdg14_cpma +
           normalized_score_sdg14_fishstocks + normalized_score_sdg14_cleanwat,
         data = LAC_data_clean)
summary(m2)

# Compare models using AIC
AIC(m1, m2)

# Compare models using ANOVA (nested comparison)
anova(m1, m2)

```

## Results

### Model 1:

1.  **Coefficients**:

-   **Intercept:** 69.02 (p-value: 0.000225)\
    The intercept is significant, indicating the baseline value for `normalized_score_sdg14_trawl` when both predictors (`normalized_score_sdg14_discard` and `normalized_score_sdg14_cpma`) are zero.

-   **normalized_score_sdg14_discard:** 0.335 (p-value: 0.031)\
    This predictor is statistically significant. For each unit increase in `normalized_score_sdg14_discard`, the `normalized_score_sdg14_trawl` is expected to increase by 0.335 units, assuming other factors remain constant.

-   **normalized_score_sdg14_cpma:** -0.166 (p-value: 0.502)\
    This predictor is **not statistically significant** (p-value \> 0.05). This suggests that `normalized_score_sdg14_cpma` does not have a significant effect on `normalized_score_sdg14_trawl` after controlling for the other variables.

2.  **Model fit:**

    -   **Residual standard error:** 20.28\
        This is the standard deviation of the residuals (the differences between observed and predicted values), indicating how much variability there is around the fitted regression line.

    <!-- -->

    -   **Multiple R-squared:** 0.2918\
        This indicates that about **29%** of the variance in `normalized_score_sdg14_trawl` is explained by `normalized_score_sdg14_discard` and `normalized_score_sdg14_cpma`.

    <!-- -->

    -   **Adjusted R-squared:** 0.1907\
        This is a more conservative estimate of model fit, adjusting for the number of predictors. It suggests that about **19%** of the variance in `normalized_score_sdg14_trawl` is explained by the model.

    <!-- -->

    -   **F-statistic:** 2.885 (p-value: 0.08931)\
        The F-statistic tests if the model is a good fit. The p-value is greater than 0.05, indicating that the model is not statistically significant at the 5% level. This means the model does not explain the data well.

### Model 2:

1.  **Coefficients:**

    -   **Intercept:** 32.29 (p-value: 0.2788)\
        The intercept is not statistically significant, meaning it doesn't provide meaningful information in this context.

    <!-- -->

    -   **normalized_score_sdg14_discard:** 0.371 (p-value: 0.0122)\
        This predictor is statistically significant. For each unit increase in `normalized_score_sdg14_discard`, the `normalized_score_sdg14_trawl` is expected to increase by 0.371 units, assuming other factors remain constant.

    <!-- -->

    -   **normalized_score_sdg14_cpma:** -0.330 (p-value: 0.1625)\
        This predictor is **not statistically significant** (p-value \> 0.05), suggesting that `normalized_score_sdg14_cpma` does not contribute significantly to explaining `normalized_score_sdg14_trawl`.

    <!-- -->

    -   **normalized_score_sdg14_fishstocks:** 0.561 (p-value: 0.0384)\
        This predictor is statistically significant. For each unit increase in `normalized_score_sdg14_fishstocks`, the `normalized_score_sdg14_trawl` is expected to increase by 0.561 units, assuming other factors remain constant.

    <!-- -->

    -   **normalized_score_sdg14_cleanwat:** 0.058 (p-value: 0.864)\
        This predictor is **not statistically significant**. This suggests that `normalized_score_sdg14_cleanwat` does not have a meaningful effect on `normalized_score_sdg14_trawl`.

2.  **Model fit:**

    -   **Residual standard error:** 17.83\
        This is a slightly lower residual standard error compared to Model 1, suggesting that Model 2 fits the data slightly better.

    <!-- -->

    -   **Multiple R-squared:** 0.5306\
        This model explains **53%** of the variance in `normalized_score_sdg14_trawl`, which is a much better fit than Model 1 (29%).

    <!-- -->

    -   **Adjusted R-squared:** 0.3742\
        This indicates that about **37%** of the variance in `normalized_score_sdg14_trawl` is explained by this model, adjusted for the number of predictors.

    <!-- -->

    -   **F-statistic:** 3.392 (p-value: 0.04473)\
        This p-value is **significant**, meaning that Model 2 is a statistically significant model at the 5% level, and it explains the variation in `normalized_score_sdg14_trawl`.

### Model Comparison:

**AIC (Akaike Information Criterion):**

-   **Model 1 (AIC = 155.27)**

-   **Model 2 (AIC = 152.28)** The lower AIC value for Model 2 indicates that it provides a better fit with fewer penalties for the additional parameters, making it a more efficient model.

    **ANOVA Comparison:**

    -   **p-value: 0.08476** (from the F-test comparing Model 1 and Model 2)

    -   This suggests that adding the additional predictors (`normalized_score_sdg14_fishstocks` and `normalized_score_sdg14_cleanwat`) improves the model, but not significantly at the 5% level.

### Conclusion:

-   **Model 2** (with four predictors) explains more of the variance in `normalized_score_sdg14_trawl` (53%) than Model 1 (29%).

-   The **statistically significant predictors** in Model 2 are `normalized_score_sdg14_discard` and `normalized_score_sdg14_fishstocks`.

-   While Model 2 is a better fit based on AIC and R-squared, the F-test and p-value suggest that the improvement is marginal.

Predicted vs Actual Scores for Trawling

### What Are Normalized Predictors?

-   **Normalized predictors** are values that have been **scaled to a common range**, often **0 to 100**.

-   This helps when comparing countries with different raw values (e.g., a small island vs. Brazil).

-   Normalization removes bias from different units (like tons of fish vs. square km of protected ocean).

-   Example: “Normalized Score: sdg14_fishstocks” = a country’s performance relative to others on fish stock sustainability.

## Predicted vs. Actual Plot

```{r}
# Predicted vs. actual for Model 2
LAC_data_clean$predicted_trawl <- predict(m2)

ggplot(LAC_data_clean, aes(x = predicted_trawl, y = normalized_score_sdg14_trawl)) +
  geom_point(size = 3, color = "#0072B2") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  labs(
    title = "Predicted vs. Actual SDG 14 Trawling Scores",
    x = "Predicted Trawling Score",
    y = "Actual Trawling Score"
  ) +
  theme_minimal()

```

### What this shows:

-   Each point is a country in the LAC region.

-   The closer a point is to the dashed line, the more accurate the prediction.

-   Helps us **see how well Model 2 performs.**
